---
fontsize: 12pt
geometry: margin=0.50in
subparagraph: yes
params: 
    title: "title"
    date:  "0000-00-00"
title: "`r params$title`"
date:  "`r params$date`"
author: "John K. Everett, Ph.D."
header-includes:
   - \usepackage{longtable}
   - \usepackage[singlelinecheck=false]{caption}
   - \usepackage{titlesec}
   - \usepackage{array}
   - \usepackage{makecell}
   - \usepackage{booktabs}
   - \usepackage{float}
   - \usepackage{pdflscape}
   - \newcommand{\blandscape}{\begin{landscape}}
   - \newcommand{\elandscape}{\end{landscape}}
output:
  pdf_document:
    toc: false
    keep_tex: no
    latex_engine: xelatex
---
\fontsize{12}{16}\selectfont
\captionsetup[table]{labelformat=empty}

\begin{center}
https://github.com/helixscript/myGOI
\end{center}

```{r titlePage, echo=FALSE, message=FALSE, warning=FALSE, fig.height=8, cache=TRUE}
library(dplyr)
library(kableExtra)

tab <- data.frame(parameter = names(config), value = unlist(config))

kable(tab, format = "latex", align='l', linesep = "", booktabs = T, row.names = FALSE, longtable = TRUE, caption = '') %>%
   kable_styling(latex_options = c("hold_position"), font_size = 10)
```

\newpage

The volcano plot below depicts changes in integration frequency of genes between early 
and later time points. Integration frequency is defined as the number of unique integration 
sites near a gene divided by the total number of integration sites recovered within the early or later 
time periods. The change in integration frequency is defined as (f - fo) / fo where f is the frequency 
during later time points and fo is the frequency during earlier time points. The significance 
of enrichment or depletion of integration events for each gene is assessed using Fisher's Exact 
tests. This plot does not correct for multiple comparisons.

\vspace{0.50cm}

```{r uncorrectedVolcanoPlot, echo=FALSE, message=FALSE, warning=FALSE, fig.height=8, cache=TRUE}
pValVolcanoPlot
```

\newpage

Alternatively, the volcano plot can be drawn where the p-values from the multiple Fisher's Exact tests 
are corrected for multiple comparisons using the Benjamini & Hochberg method.

\vspace{0.50cm}

```{r, correctedVolcanoPlot, echo=FALSE, message=FALSE, warning=FALSE, fig.height=8, cache=TRUE}
pValAdjVolcanoPlot
```

\newpage

Genes of interest can be associated with four possible categories:
```{r, DEAL_table, echo=FALSE, message=FALSE, warning=FALSE, fig.height=8, cache=TRUE}
tab <- t(tibble('Depleted (D)' = 'Depleted genes show a significant decrease (p < 0.05) in integration frequency at later time points.',
                'Enriched (E)' = 'Enriched genes show a significant increase (p < 0.05) in integration frequency at later time points.',
                'Abundant (A)' = 'Abundant genes are genes associated with the top 1% of clonal abundance estimates at later time points.',
                'Longitudinal (L)' = paste0('Longitudinal genes are genes with at least 3 unique integrations in at least ', 
                                          config$longitudinal_minNumSubjects, ' subjects recovered after ', config$longitudinal_minTimeDays, ' days.')))        

kable(tab, format = "latex", align='l', linesep = "", booktabs = T, row.names = TRUE, longtable = TRUE, caption = '') %>%
   kable_styling(font_size = 9)   
```

The assignment of genes to the Enriched and Depleted categories is dependent on gene specific Fisher's Exact tests (uncorrected p-values $\leq$ 0.05). 
Genes associated with the Enriched, Abundant, and Longitudinal categories are of particular interest.  
  
\vspace{0.5cm}

```{r, DEAL_legend, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
tab <- t(tibble('gene' = 'Gene symbol.',
                'subjects' = 'Total number of subjects with an integration near gene.',
                'percentChange' = 'Percent increase in integration frequency compared to earlier time period.',
                'maxAbund' =  paste0('Max. abundance estimate observed > ', config$earlyVsLateCutoffDays, ' days.'),
                'longitudinalSites' = paste0('Number of integration sites observed >= ', config$longitudinal_minTimeDays, ' days.'),
                'categories' = 'DEAL categories associated with gene.'))

cat(kable(tab, format = "latex", align='l', linesep = "", booktabs = T, row.names = TRUE) %>%
   kable_styling(latex_options = c("hold_position"), font_size = 10))
```


```{r, ELA_genes,echo=FALSE, message=FALSE, warning=FALSE}
tab <- select(subset(k, flag == 'EAL'), gene, subjects, totalSites, percentChange, maxAbund, longitudinalSites, flag)
tab <- arrange(tab, desc(percentChange))
tab$percentChange <- sprintf("%.1f%%", tab$percentChange)
tab <- dplyr::rename(tab, 'categories' = flag)

kable(tab,
      format = "latex", linesep = "", align='c', booktabs = T, row.names = FALSE, caption = 'Table 1. Genes annotated as EAL.') %>%
      kable_styling(latex_options = c("hold_position"))
```

  
\newpage

Genes associated with the Enriched category have greater integration frequencies at later time points compared to earlier time points
which suggests that integration near these genes bolsters cell survival.

\vspace{0.25cm}

```{r Table2_legend, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
tab <- t(tibble('gene' = 'Gene symbol.',
                'subjects' = 'Total number of subjects with an integration near gene.',
                'earlyCount' = paste0('Number of integration sites recovered from earlier time points (<= ', config$earlyVsLateCutoffDays, ' days).'),
                'lateCount' = paste0('Number of integration sites recovered from later time points (> ', config$earlyVsLateCutoffDays, ' days).'),
                'percentChange' = paste0('Percent increase in integration frequency compared to earlier time period (> ', config$earlyVsLateCutoffDays, ' days).'),
                'pVal' = "p-value from Fisher' Exact test.",
                'categories' = 'DEAL categories associated with gene.'))

kable(tab, format = "latex", align='l', linesep = "", booktabs = T, row.names = TRUE) %>%
   kable_styling(latex_options = c("hold_position"), font_size = 10)
```

```{r, E_table, echo=FALSE, message=FALSE, warning=FALSE}
caption <- 'Table 2. Top 100 enriched genes.'
if(nrow(k[grepl('E', k$flag),]) <= 100) caption <- 'Table 2. Enriched genes.'

tab <- select(k[grepl('E', k$flag),], gene, subjects, earlyCount, lateCount, percentChange, pVal, flag) %>% arrange(pVal, desc(percentChange)) %>% dplyr::slice(1:100)
tab$percentChange <- sprintf("%.2f%%", tab$percentChange)
tab$pVal <- formatC(tab$pVal, format = "e", digits = 2)
tab <- dplyr::rename(tab, 'categories' = flag)

kable(tab, format = "latex", align='c', linesep = "", booktabs = T, row.names = FALSE, longtable = TRUE, caption = caption) %>%
   kable_styling(latex_options = c("scale_down", "hold_position", "repeat_header"), font_size = 9) 
```

\newpage

Genes associated with the Depleted category have lower integration frequencies at later time points compared to earlier time points
which suggests that integration near these genes may be detrimental to cell survival.

\vspace{0.25cm}

```{r Table3_legend, echo=FALSE, message=FALSE, warning=FALSE}
tab <- t(tibble('gene' = 'Gene symbol.',
                'subjects' = 'Total number of subjects with an integration near gene.',
                'earlyCount' = paste0('Number of integration sites recovered from earlier time points (<= ', config$earlyVsLateCutoffDays, ' days).'),
                'lateCount' = paste0('Number of integration sites recovered from later time points (> ', config$earlyVsLateCutoffDays, ' days).'),
                'percentChange' = paste0('Percent increase in integration frequency compared to earlier time period (> ', config$earlyVsLateCutoffDays, ' days).'),
                'pVal' = "p-value from Fisher' Exact test.",
                'categories' = 'DEAL categories associated with gene.'))

kable(tab, format = "latex", align='l', linesep = "", booktabs = T, row.names = TRUE) %>%
   kable_styling(latex_options = c("hold_position"), font_size = 10)
```

```{r, D_table, echo=FALSE, message=FALSE, warning=FALSE}
caption <- 'Table 3. Top 100 depleated genes.'
if(nrow(k[grepl('D', k$flag),]) <= 100) caption <- 'Table 3. Depleated genes.'

tab <- select(k[grepl('D', k$flag),], gene, subjects, earlyCount, lateCount, percentChange, pVal, flag) %>% arrange(pVal, percentChange) %>% dplyr::slice(1:100)
tab$percentChange <- sprintf("%.2f%%", tab$percentChange)
tab$pVal <- formatC(tab$pVal, format = "e", digits = 2)
tab <- dplyr::rename(tab, 'categories' = flag)

kable(tab, format = "latex", align='c', linesep = "", booktabs = T, row.names = FALSE, longtable = TRUE, caption = caption) %>%
   kable_styling(latex_options = c("scale_down", "hold_position", "repeat_header"), font_size = 9) 
```
  
\newpage

Genes associated with the Abundant category reached high levels of clonal abundance measured by the sonic abundance method
at later time points which suggests that integration near these genes may bolster cell division. 

\vspace{0.25cm}

```{r Table4_legend, echo=FALSE, message=FALSE, warning=FALSE}
tab <- t(tibble('gene' = 'Gene symbol.',
                'subjects' = 'Total number of subjects with an integration near gene.',
                'lateCount' = paste0('Number of integration sites recovered after ', config$earlyVsLateCutoffDays, ' days.'),
                'maxAbund' = paste0('Maximum estimated clonal abundance observered after ', config$earlyVsLateCutoffDays, ' days.'),
                'maxRelAbund' = paste0('Maximum relative sample clonal abundance observered after ', config$earlyVsLateCutoffDays, ' days.'),
                'categories' = 'DEAL categories associated with gene.'))

kable(tab, format = "latex", align='l', linesep = "", booktabs = T, row.names = TRUE) %>%
   kable_styling(latex_options = c("hold_position"), font_size = 10)
```

```{r, A_table, echo=FALSE, message=FALSE, warning=FALSE}
caption <- 'Table 4. Top 100 abundant genes.'
if(nrow(k[grepl('A', k$flag),]) <= 100) caption <- 'Table 4. Abundant genes.'

tab <- select(k[grepl('A', k$flag),], gene, subjects, lateCount, maxAbund, maxRelAbund, flag) %>% arrange(desc(maxAbund)) %>% dplyr::slice(1:100)
tab$maxRelAbund <- sprintf("%.2f%%", tab$maxRelAbund)
tab <- dplyr::rename(tab, 'categories' = flag)


kable(tab, format = "latex", align='c', linesep = "", booktabs = T, row.names = FALSE, longtable = TRUE, caption = caption) %>%
   kable_styling(latex_options = c("scale_down", "hold_position", "repeat_header"), font_size = 9) 
```
  
\newpage

Genes are categorized as Longitudinal if $\geq$ `r config$longitudinal_minNumSubjects` different integrations across at least 
`r config$longitudinal_minNumSubjects` different subjects are observed $\geq$ `r config$longitudinal_minTimeDays` days post-transduction.

\vspace{0.25cm}

```{r, L_legend, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
leg <- t(tibble('gene' = 'Gene symbol.',
                'totalSites' = 'total number of unique integrations associated with gene.',  
                'longitudinalSites' = paste0('total number of unique integrations associated with gene recovered >= ', 
                                           config$longitudinal_minTimeDays, ' days.'),  
                'longitudinalSubjects' =  paste0('number of subjects associated with integrations associated with gene recovered >= ', 
                                           config$longitudinal_minTimeDays, ' days.'),    
                'longitudinalTimePoints' =  paste0('number of time points sampled >= ', 
                                           config$longitudinal_minTimeDays, ' days.'),    
                'latestTimePointDays' = 'last time point sampled containing integrations in gene.'))  

kable(leg, format = "latex", align='l', linesep = "", booktabs = T, row.names = TRUE, longtable = TRUE, caption = '') %>%
   kable_styling(latex_options = c("hold_position"), font_size = 10)
```

```{r, L_table, echo=FALSE, message=FALSE, warning=FALSE}
caption <- 'Table 5. Top 100 longitudinaly persistant genes.'
if(nrow(k[grepl('L', k$flag),]) <= 100) caption <- 'Table 5. Longitudinaly persistant genes.'

tab <- select(k[grepl('L', k$flag),], gene, totalSites, longitudinalSites, longitudinalSubjects, longitudinalTimePoints, latestTimePointDays) %>% 
  arrange(desc(latestTimePointDays), desc(longitudinalSites)) %>% dplyr::slice(1:100)


kable(tab, format = "latex", align='c', linesep = "", booktabs = T, row.names = FALSE, longtable = TRUE, caption = caption) %>%
   kable_styling(latex_options = c("scale_down", "hold_position", "repeat_header"), font_size = 9) 
```
  
